---
title: "MLM Mini Project"
date:  "`r format(Sys.time(), '%B %d %Y')`"
output:
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
###Load libraries
if(!requireNamespace("here"))
  install.packages("here", repos = "https://cloud.r-project.org")
require("here")

if(!requireNamespace("lme4"))
  install.packages("lme4", repos = "https://cloud.r-project.org")
require("lme4")

if(!requireNamespace("lmerTest"))
  install.packages("lmerTest", repos = "https://cloud.r-project.org")
require("lmerTest")

options(scipen = 999)
```

## Team Members and division of work: 

```{r}
# Insert code to set.seed
set.seed(2042001)
```


### Question 1:
You will generate simulated data for a single school with 100 classrooms, each of which has 200 students. 

  a. Outcome for student $i$ in classroom $j$: $Y_{ij}$.
    
  b. There is a single predictor, $X_{ij} \sim U(0,1)$ (uniform on [0,1])
    
  c. There is a classroom random effect, $\eta_j \sim N(0,\sigma^2_{\eta})$, where $\sigma^2_{\eta} = 2$. 
  
  d. Subject level error, $\varepsilon_{ij} \sim N(0, \sigma^2_{\varepsilon})$, where  $\sigma^2_{\varepsilon} = 2$. 
  
  e. `set.seed(2042001)` once at the beginning of your code.
  
  f. Generate the random quantities in this order to ensure the same solution for everyone: X, $\eta_j$, $\varepsilon_{ij}$ 
  
  g. The outcome has the following form (DGP, given the modeling parameters above):
$$Y_{ij} = 0 + 1X_{ij} + \eta_j + \varepsilon_{ij} ; \mbox{     } \eta_j \sim N(0,\sigma^2_{\eta}), \varepsilon_{ij} \sim N(0, \sigma^2_{\varepsilon}), indep. $$
h. Generate a single simulated dataset (you will need a "classid" variable to track classrooms); you can optionally assign a "studentid")
  i. **Important:** construct classid such that classrooms appear consecutively within the dataframe. As per: `rep(1:J,each=n_j)`
  
  
```{r}
# Insert code to generate data and outcome variable, store variables in a dataframe

# set size assumptions
n.classrooms <- 100
n.stu.per.class <- 200

# generate data
X_ij <- runif(n.classrooms * n.stu.per.class, min = 0, max = 1)
eta_j <- rnorm(n.classrooms, mean = 0, sd = 2)
epsilon_ij <- rnorm(n.classrooms * n.stu.per.class, mean = 0, sd = 2)

# calculate outcome variable REVIEW THIS DGP !!!!!!!!!!!
Y_ij <- 0 + 1*X_ij + rep(eta_j, each = n.stu.per.class) + epsilon_ij

# store variables in dataframe
dat <- data.frame(
  studentid = 1:(n.classrooms * n.stu.per.class),
  classid = rep(1:n.stu.per.class, each = n.classrooms),
  predictor = X_ij,
  outcome = Y_ij
)
```

  
### Question 2:  REVIEW THIS MODEL !!!!!!!!!!!
Fit the model corresponding to the DGP on your simulated data.
```{r}
# Insert code to fit model and print summary
lm1 <- lmerTest::lmer(outcome ~ predictor + (1 | classid), data = dat, REML = TRUE)
summary(lm1)
```

a. Report coefficient estimate for slope on X.

    Response: 

b. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**

```{r}
# Insert code to compute confidence interval 
coefs <- summary(lm1)$coefficients
lower <- coefs[2,1] - (coefs[2,2] * 2)
upper <- coefs[2,1] + (coefs[2,2] * 2)
print(paste0("The 95% confidence band is [", round(lower, 3), ", ", round(upper, 3), "]"))
```

    Response:
    
### Question 3:
3. Next, we simulate missing data in several ways. This is the first:
a. Make a copy of the data, then modify the copy following these instructions:
```{r}
# Insert code to make a copy of the data
dat2 <- dat
```

b. Generate $Z_{ij} \sim$ Bernoulli$(p)$, with $p = 0.5$
c. Set Y to NA when $Z_{ij} == 1$. This should look a lot like "MCAR" missingness.
```{r}
# Insert code the generate your data
Z_ij <- rbinom(n = n.classrooms*n.stu.per.class, size = 1, prob = 0.5)
dat2$outcome <- ifelse(Z_ij == 1, NA, dat2$outcome)
```

d. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well.
```{r}
# Insert code to fit model and compute confidence interval
lme2 <- lmerTest::lmer(outcome ~ predictor + (1 | classid), data = dat2)
summary(lme2)

# calculate confidence band
coefs <- summary(lme2)$coefficients
lower <- coefs[2,1] - (coefs[2,2] * 2)
upper <- coefs[2,1] + (coefs[2,2] * 2)
print(paste0("The 95% confidence band is [", round(lower, 3), ", ", round(upper, 3), "]"))
```

e. Do you see any real change in the $\beta_X$ estimate? **comment**
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data?
    
    Response: 

f. What is the total sample size $N$ used in the model fit? **comment**

```{r}
# number of observations used
sum(!is.na(dat2$outcome))
length(summary(lme2)$residuals)

# number of groups used
summary(lme2)$ngrps
```

    Response:
    
### Question 4: 
Missing Data II: Make another copy of the original data, then modify the copy as follows:
a. Generate $Z_{ij} \sim$ Bernoulli$(X_{ij})$, with $X_{ij}$ your predictor generated previously.
b. Set Y to NA when $Z_{ij} == 1$. This should look a lot like "MAR" missingness.
```{r}
# Insert code the generate your data
dat3 <- dat
Z_ij <- rbinom(n = n.classrooms*n.stu.per.class, size = 1, prob = dat3$predictor)
dat3$outcome <- ifelse(Z_ij == 1, NA, dat3$outcome)
```
c. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well. **comment**
```{r}
# Insert code to fit model and compute confidence interval
lme3 <- lmerTest::lmer(outcome ~ predictor + (1 | classid), data = dat3)
summary(lme3)

# calculate confidence band
coefs <- summary(lme3)$coefficients
lower <- coefs[2,1] - (coefs[2,2] * 2)
upper <- coefs[2,1] + (coefs[2,2] * 2)
print(paste0("The 95% confidence band is [", round(lower, 3), ", ", round(upper, 3), "]"))
```

    Response: 

d. Do you see any real change in the $\beta_X$ estimate?
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**
    
    Response:

e. What is the total sample size $N$ used in the model fit? **comment**

```{r}
# number of observations used
sum(!is.na(dat3$outcome))
length(summary(lme3)$residuals)

# number of groups used
summary(lme3)$ngrps
```

    Response:

### Question 5: 
Missing Data III: Make another copy of the original data, then modify the copy as follows:
```{r}
# Insert code to make a copy of the original data
dat4 <- dat
```

a. First, define the expit function: `expit <- function(x) exp(x)/(1+exp(x))`
```{r}
# Insert code to define expit function
expit <- function(x) exp(x)/(1 + exp(x))
```

b. Generate $Z_{ij} \sim$ Bernoulli$(expit(Y_{ij}))$, with $Y_{ij}$ your *outcome* generated previously. 

c. Set Y to NA when $Z_{ij} == 1$. This should look like a violation of "MAR" missingness (missingness depedents on outcome and cannot be *simply* predicted with the predictor set -- Y should be correlated with X, though, so it might not be too bad a violation). 
```{r}
# Insert code the generate your data
Z_ij <- rbinom(n = n.classrooms*n.stu.per.class, size = 1, prob = expit(dat4$outcome))
dat4$outcome <- ifelse(Z_ij == 1, NA, dat4$outcome)
```

d. Refit the model on the new data and report the coefficient estimate for slope on X. Look at the other parameter estimates as well. **comment**
```{r}
# Insert code to fit model and compute confidence interval
lme4 <- lmerTest::lmer(outcome ~ predictor + (1 | classid), data = dat4)
summary(lme4)

# calculate confidence band
coefs <- summary(lme4)$coefficients
lower <- coefs[2,1] - (coefs[2,2] * 2)
upper <- coefs[2,1] + (coefs[2,2] * 2)
print(paste0("The 95% confidence band is [", round(lower, 3), ", ", round(upper, 3), "]"))
```

    Response:

e. Do you see any real change in the $\beta_X$ estimate? **comment**
    i. Does a 95% confidence band for this coefficient estimate cover the “truth” that you used to generate the data? **comment**
    
      Response:
    
f. What is the total sample size $N$ used in the model fit? **comment**

```{r}
# number of observations used
sum(!is.na(dat4$outcome))
length(summary(lme4)$residuals)

# number of groups used
summary(lme4)$ngrps
```

    Response:







